p-adic stability in linear algebra

### 2015-04-20 examples/stability.tex
Updates for final version

\newcommand{\noopsort}[1]{}
\DeclareMathOperator{\NP}{NP}
\DeclareMathOperator{\HP}{HP}
\DeclareMathOperator{\PP}{PP}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\End}{End}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\val}{val}
\DeclareMathOperator{\pr}{pr}
\DeclareMathOperator{\tr}{Tr}
\DeclareMathOperator{\com}{Com}
\DeclareMathOperator{\Grass}{Grass}
\DeclareMathOperator{\Lat}{Lat}
\DeclareMathOperator{\round}{round}

@@

@@

\vspace{1mm}
 \noindent
 {\bf Categories and Subject Descriptors:} \\
\noindent I.1.2 [{\bf Computing Methodologies}]:{~} Symbolic and Algebraic
  Manipulation -- \emph{Algebraic Algorithms}

 \vspace{1mm}
 \noindent
 {\bf General Terms:} Algorithms, Theory

 \vspace{1mm}
 \noindent
 {\bf Keywords:} $p$-adic precision, linear algebra, ultrametric analysis
\medskip

@@

In \cite{caruso-roe-vaccon:14a}, we advocate the use of lattices
to track the precision of vectors and points on $p$-adic manifolds.
The main result of \emph{loc.~cit.}, recalled here as Proposition \ref{prop:precision},
allows for the propagation of precision using differentials.  While
allowing arbitrary lattices in representing precision carries a high
cost in space and time requirements, the reduced precision loss
can sometimes make up for these costs as part of a larger algorithm.
In this paper we apply the ideas of \cite{caruso-roe-vaccon:14a} to
a variety of linear algebraic tasks.

@@

In Section \ref{ssec:mulmatrix}, we analyze matrix multiplication and report
on experiments that demonstrate the utility of lattice-based precision tracking.
Section \ref{ssec:det} finds the precision of the determinant of a
matrix, and Section \ref{ssec:charpoly} applies the resulting formula to
characteristic polynomials.  We define the precision polygon of a matrix, which
gives a lower bound on the precision of the characteristic polynomial.
The precision polygon lies above the Hodge polygon of the matrix, and
we give statistics on the difference and the amount of diffuse precision.
Finally, we find the precision of LU factorization, determine the conditions under which
Proposition \ref{prop:precision} may be applied, and again describe experiments
with lattice-based precision analysis.

@@

\noindent
{\bf Notation.}
Throughout the paper, the letter $K$ will refer to a complete discrete 
valuation field. Usual examples are finite extensions of $\Qp$ and 
Laurent series fields over a field. We denote by $\val : K \to \Z \cup 
\{+\infty\}$ the valuation on $K$ normalized by $\val(K^\star) = \Z$, by 
$\OK$ the ring of integers and by $\pi \in K$ an element of valuation 
$1$. We let $\Vert \cdot \Vert$ be the norm associated to $\val$.

@@

Roughly speaking, the main proposal of \cite{caruso-roe-vaccon:14a} is 
to use lattices to track precision when computing with elements lying in 
vector spaces over $K$.

More explicitly, let us consider a finite 
dimensional\footnote{The framework of \cite{caruso-roe-vaccon:14a} is 
actually those of Banach spaces. However, we will not need infinite 
dimensional spaces in the present paper.} normed vector space $E$ 
defined over $K$. We use the notation $\Vert \cdot \Vert_E$ for the norm 
on $E$ and $B^-_E(r)$ (resp. $B^{\phantom -}_E(r)$) for the open (resp. 
closed) ball of radius $r$ centered at the origin. A lattice $L \subset 
E$ is, by definition, a sub-$\O_K$-module which generates $E$ over $K$. 
Since we are working in a ultrametric world, the balls $B^{\phantom 
-}_E(r)$ and $B^-_E(r)$ are examples of lattices. Actually, lattices 
should be thought of as special neighborhoods of $0$ and therefore are good 
candidates to model precision data. Moreover, as revealed in 
\cite{caruso-roe-vaccon:14a}, they behave quite well under (strictly) 
differentiable maps:

@@

Proposition \ref{prop:precision} can be rephrased in terms of precision 
as follows: if the input $v_0$ is known at precision $H$, then $f(v_0)$ 
is known up to $f'(v_0)(H)$ and this precision is optimal. Optimality is 
reflected by the equality sign in Eq.~\eqref{eq:firstorder}.

In \cite{caruso-roe-vaccon:14a}, we also explained that if the function 
$f$ is locally analytic, then the constant $\delta$ appearing in
Proposition \ref{prop:precision} can be made explicit in terms of the
growing function $\Lambda(f)$ of $f$ defined by:
$$\textstyle \Lambda(f)(v) = 
\log \big( \sup_{h \in B^-_E(e^v)} \Vert 
f(h) \Vert \big)$$
with the convention that $\Lambda(f)(v) = +\infty$ if $f$ does not
converge on $B^-_E(e^v)$.
We refer to \cite[Proposition 3.12]{caruso-roe-vaccon:14a} for the
precise statement.  We state here the case of
integral polynomial functions. A function $f : E \to F$ is said to be
\emph{integral polynomial} if it is given by multivariate polynomial 
functions with coefficients in $\O_K$ in any (equivalently all) system 
of coordinates associated to a $\OK$-basis of $B_E(1)$.

@@

In \cite[Appendix A]{caruso-roe-vaccon:14a}, the theory is extended 
to manifolds over $K$ and covers in this way a very large family of
situations (\emph{e.g.} computation with vector spaces as discussed
in \S \ref{sec:vectorspaces}). In this extended framework, the precision 
datum at some point $x$ is a lattice in the tangent space at $x$. 
Propositions \ref{prop:precision} and \ref{prop:precision2} have
analogues as well, simply obtained by working in charts.

In this paper, we will compare the use of lattices to track precision
with the standard coordinate-wise methods.  The following definition
will be useful in discussing the differences.  Suppose $E$ is equipped
with a basis $(e_1, \ldots, e_n)$ and write $\pi_i : E \to Ke_i$ for the projections.

\begin{deftn} \label{def:diffused}
If $H \subset E$ is a lattice, let
\[
H_0 = \pi_1(H) \oplus \cdots \oplus \pi_n(H).
\]
Set $k = \text{length}(H_0/H)$.  Then we say that $H$ has $k$
\emph{diffused digits of precision}.
\end{deftn}
If $H$ represents the actual precision of some object, then
$H_0$ is the best coordinate-wise approximation to $H$.
The diffused precision is not visible coordinate-wise in the
specified basis of $E$.

@@

We assume that we are given in addition two nondecreasing convex 
functions $\Lambda_g$ and $\Lambda_h$ such that $\Lambda(g) \leq 
\Lambda_g$ and $\Lambda(h) \leq \Lambda_h$. We suppose further that 
there exists $\nu$ such that $\Lambda_g$ is constant on the interval 
$({-}\infty, \nu]$\footnote{We note that this assumption is fulfilled if 
we take $\Lambda_g = \Lambda(g)$ because we have assumed that $g(0)$ 
does not vanish.}. We introduce the functions $\tau_\nu$ and $\Lambda_f$ 
defined by:
$$\begin{array}{rll}
&\tau_\nu(x) = x & \text{if } x \leq \nu \\
& \hphantom{\tau_\nu(x)} = {+} \infty & \text{otherwise} \medskip \\
\text{and} &
\multicolumn{2}{l}{\Lambda_f(x) = 
  \tau_\nu \circ (\id + \Lambda_g \circ \Lambda_h)(x + \alpha)}
\end{array}$$
where $\alpha$ is a real number satisfying $\Vert n! \Vert \geq 
e^{-\alpha n}$ for all $n$. If $p$ is the characteristic of the residue field,
a suitable value for $\alpha$ is $\alpha = - 
\frac p {p-1} \cdot \log \Vert p \Vert$ if $p > 0$ and $\alpha = 0$ if $p = 0$.
The next Proposition is proved in Appendix \ref{app:proof}.

@@

@@

The computation of the differential of $\det : M_{n,n}(K) \to K$
is classical: at a point $M$ it is the linear map
\[
\detp(M) : dM \mapsto \tr(\com(M) \cdot dM),
\]
where $\com(M)$ stands for the comatrix of $M$, which is $\det(M) M^{-1}$ when $M$ is invertible. If $M$ is not 
invertible, $\detp(M)$ is not surjective and we cannot apply Proposition 
\ref{prop:precision}. Therefore, we suppose that $M$ is invertible for the rest of this section.

@@

Recall that the \emph{Newton polygon} $\NP(f)$ of a polynomial $f(X) = 
\sum_{i} a_i X^i$ is the lower convex hull of the points $(i, 
\val(a_i))$ and the \emph{Newton polygon} $\NP(M)$ of a matrix $M$ is 
$\NP(\charpoly(M))$.  The \emph{Hodge polygon} of $M$ is the lower 
convex hull of the points $(i, \sum_{j = 1}^{n-i} \sigma_j(M))$. For 
any matrix $M$, the Newton polygon lies above the Hodge polygon 
\cite{kedlaya:padicDiffEq}*{Thm. 4.3.11}.

Such polygons arise naturally in tracking the precision of polynomials 
\cite{caruso-roe-vaccon:14a}*{\S 4.2}.  Any such polygon $P$ yields a 
lattice $\mathcal{L}_P$ in $K_{<n}[X]$ consisting of polynomials whose 
Newton polygons lie above $P$.  This lattice is generated by monomials 
$a_iX^i$, where $\val(a_i)$ is the ceiling of the height of $P$ at $i$.  
One uses these polygons to do a certain kind of coordinate-wise 
precision tracking for polynomials, bounding the actual precision 
lattice. We now introduce another polygon, bounded 
between these two, that will provide an estimate on the precision of $\charpoly(M)$.

@@

It is clear from the definition $\mathcal L_{\PP(M)} \subset 
\charp(M)(\mathcal{L}_0)$ where $\mathcal{L}_0$ is the standard lattice 
$M_{n,n}(\OK)$. More precisely $\PP(M)$ is the smallest polygon $P$ for 
which the inclusion $\mathcal L_P \subset \charp(M)(\mathcal{L}_0)$ holds. By 
Proposition \ref{prop:precision}, the precision polygon determines the minimal 
precision losses possible when encoding polynomial precision using polygons.

@@

It remains to prove the comparison with Newton polygon.
Set $f = \charpoly(M)$, set $m_{i,j}$ as the $(i,j)$-th entry of $M$, 
$f_{i,j}$ as the $(i,j)$-th entry of $\com(X{-}M)$ and $\mu_{i,j} = 
\val(m_{i,j})$. We write $f[k]$ 
for valuation of the coefficient of $X^k$ in $f$, and set $f[-1] = 
+\infty$. The equation $(X{-}M) \cdot \com(X{-}M) = f \cdot I$ yields, 
for all $i$ and $k$,
\begin{align*}
f[k] &\ge \inf(f_{i,i}[k{-}1], \mu_{i,0} + f_{0,i}[k], \ldots, 
\mu_{i,n} + f_{n,i}[k]) \\
&\ge \inf(f_{i,i}[k{-}1], f_{j,i}[k]),
\end{align*}
with the infimum over $j$. 
Taking lower convex hulls and noting that $\PP(M)$ is nonincreasing,
which follows from the comparison with the Hodge polygon, we get the 
result.
\end{proof}

@@

In conclusion, for randomly chosen matrices, approximating $\mathcal{E}$ 
using the Hodge polygon loses only a small amount of precision. However,
if the $\sigma_i(M)$'s are known to be large or if $M$ is a translate of
such a matrix, using lattice precision can be very useful.

@@

%This example has been previously studied by Caruso in \cite{caruso:12a}, 
%but without using differential precision. We are going to show here that 
%the framework of \S \ref{ssec:boundLambdaf} applies directly.
In this section, we denote by $\Vert \cdot \Vert$ the subordinate matrix norm on $M_n(K)$
and, given a positive real number $C$, we let $B(C)$ be the closed ball 
in $M_n(K)$ centered at the origin of radius $C$. We consider in addition
the following subsets of $M_n(K)$:

@@

\noindent
{\bf Numerical experiments.}
Let $B_n=(E_{i,j})_{1 \leq i,j\leq n}$ be the canonical basis of 
$M_n(K)$. It can be naturally seen as a basis of $L_n^0 \times U_n$ as
well. For a given $M \in O_n$, let us abuse notation and write $D'(M)$
for the matrix of this linear mapping in the above basis. We remark that
$D'(M)$ is lower-triangular in this basis. Projecting $D'(M)$ onto each
coordinate, we find the best coordinate-wise loss of precision we can 
hope for the computation of $D$ is given by $\sum_u \left( \max_v \left( 
\val(D'(M)_{u,v}) \right) \right)$. This number should be compared to
$\val(\det(D'(M)))$, which is the lattice-method loss of precision. 
%With the language of Definition \ref{def:diffused},
The number of diffused digits of precision of $D'(M)(M_{n,n}(\OK))$ is 
then the difference between these two numbers.
Figure \ref{fig:LU} summarizes the mean and standard deviation of those 
losses for a sample of 2000 random matrices in $M_{d,d}(\mathbb{Z}_2)$.
%
\begin{figure}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|c|c|c|c|c|}
\hline 
& \multicolumn{4}{|c|}{Loss of precision in LU decomposition} \\
\cline{2-5}
\raisebox{0.2em}{matrix}
& \multicolumn{2}{|c|}{coord-wise method} & \multicolumn{2}{|c|}{lattice method}  \\  \cline{2-5}
\smash{\raisebox{0.6em}{size}} 
& \hspace{0.5em}mean\hspace{0.5em} & deviation 
& \hspace{0.5em}mean\hspace{0.5em} & deviation \\ \hline
$2$ & $\phantom{2}3.0$& $5$ & $1.5$ & $1.4$ \\% \hline
$3$& $\phantom{2}9.4$& $11$ & $2.3$ & $2.3$\\% \hline
$4$ & $20\phantom{.4}$& $20$ & $3.8$ & $3.1$\\ \hline
\end{tabular} 
\smallskip

@@

\noindent
{\bf Inverse images.}
We now consider the inverse image mapping $\II$ sending a 
pair $(f,W) \in \mathcal W = \Hom(E,F) \times \Grass(F,d)$ to 
$f^{-1}(W)$. As before, this map does not take its values in a single 
Grassmannian and we need to stratify $\mathcal W$ in order to get 
differentiable functions. For each integer $r \in [0,n]$, we introduce 
the submanifold $\mathcal W_s$ of $\mathcal W$ consisting of those 
pairs $(f,W)$ such that $\dim f^{-1}(W) = s$. For all $s$, $\II$ 
induces a continuous function
$\II_r : \mathcal W_r \to \Grass(E,s)$.
Pick $(f,W) \in \mathcal W_r$. Set $V = f^{-1}(W)$ and denote by $w : F 
\to F/W$ the canonical projection.
Similarly to what we have done for direct images, one can prove that
the tangent space of $\mathcal W_r$ at some point $(f,W) \in \mathcal
W_r$ is the subspace of $\Hom(E,F) \times \Hom(W,F/W)$ consisting of 
pairs $(df,dw)$ such that
$d \tilde v = (w \circ df + dw \circ f)_{|W}$
factors through the linear mapping $\varphi : E/V \to F/W$ induced by 
$f$. Furthermore $\II_r$ is differentiable at $(f,W)$ and its 
differential is the linear mapping that takes $(df,dw)$ as above to the 
unique element $dv \in \Hom(V,E/V)$ satisfying $\varphi \circ dv =
d\tilde v$.

Direct images and inverse images are related by duality
as follows: if $f : E \to F$ is any linear map and $W$ is a subspace
of $F$, then $f^\star(W^\perp) = f^{-1}(W)^\perp$. We 
can thus deduce the differentials of $\DI_r$ from those of $\II_s$ and 
\emph{vice versa}.

@@

\noindent
{\bf Sums and intersections.}
Let $d_1$ and $d_2$ be two nonnegative integers. We consider the 
function $\Sigma$ defined on the manifold $\mathcal C = \Grass(E,d_1) 
\times \Grass(E,d_2)$ by $\Sigma(V_1, V_2) = V_1 + V_2$. As before, in 
order to study $\Sigma$, we stratify $\mathcal C$ according to the 
dimension of the sum: for each integer $d \in [0, d_1+d_2]$, we define 
$\mathcal C_d$ as the submanifold of $\mathcal C$ consisting of those 
pairs $(V_1, V_2)$ such that $\dim(V_1 + V_2) = d$. We get a 
well-defined mapping $\mathcal C_d \to \Grass(E,d)$ whose differential
can be computed as before. The
tangent space of $\mathcal C_d$ at a given point $(V_1, V_2)$ consists 
of pairs $(dv_1, dv_2) \in \Hom(V_1, E/V_1)
\times \Hom(V_2, E/V_2)$ such that $dv_1 \equiv dv_2 \pmod{V_1 + V_2}$ 
on the intersection $V_1 \cap V_2$ and the differential of $\Sigma_r$
at $(V_1, V_2)$ maps $(dv_1, dv_2)$ to $dv \in \Hom(V, E/V)$ (with $V
= V_1 + V_2$) defined by $dv(v_1 + v_2) = dv_1(v_1) + dv_2(v_2)$ ($v_1
\in V_1$, $v_2 \in V_2$).

>>>>

\newcommand{\noopsort}[1]{}
\DeclareMathOperator{\NP}{NP}
\DeclareMathOperator{\HP}{HP}
\DeclareMathOperator{\PP}{PP}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\End}{End}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\val}{val}
\DeclareMathOperator{\pr}{pr}
\DeclareMathOperator{\tr}{Tr}
\DeclareMathOperator{\com}{Com}
\DeclareMathOperator{\Grass}{Grass}
\DeclareMathOperator{\Lat}{Lat}
\DeclareMathOperator{\round}{round}
\DeclareMathOperator{\rank}{rank}

@@

\conferenceinfo{ISSAC'15,}{July 6--9, 2015, Bath, United Kingdom.}
\copyrightetc{Copyright \copyright~2015 ACM \the\acmcopyr}
\crdata{978-1-4503-3435-8/15/07}

@@

\category{I.1.2}{Computing Methodologies}{Symbolic and Algebraic Manipulation -- \emph{Algebraic Algorithms}}
\terms{Algorithms, Theory}
\keywords{$p$-adic precision, linear algebra, ultrametric analysis}

%\vspace{1mm}
% \noindent
% {\bf Categories and Subject Descriptors:} \\
%\noindent I.1.2 [{\bf Computing Methodologies}]:{~} Symbolic and Algebraic
%  Manipulation -- \emph{Algebraic Algorithms}
%
% \vspace{1mm}
% \noindent
% {\bf General Terms:} Algorithms, Theory
%
% \vspace{1mm}
% \noindent
% {\bf Keywords:} $p$-adic precision, linear algebra, ultrametric analysis
%\medskip

@@

In \cite{caruso-roe-vaccon:14a}, we advocate the use of lattices
to track the precision of vectors and points on $p$-adic manifolds.
The main result of \emph{loc.~cit.}, recalled in Proposition \ref{prop:precision},
allows for the propagation of precision using differentials.  While
representing precision by lattices carries a high
cost in space and time requirements, the reduced precision loss
can sometimes overwhelm these costs within a larger algorithm.
In this paper, we apply the ideas of \cite{caruso-roe-vaccon:14a} to
certain linear algebraic tasks.

@@

In Section \ref{ssec:mulmatrix}, we analyze matrix multiplication and report
on experiments that demonstrate the utility of lattice-based precision tracking.
Section \ref{ssec:det} finds the precision of the determinant of a
matrix, and Section \ref{ssec:charpoly} applies the resulting formula to
characteristic polynomials.  We define the precision polygon of a matrix, which
gives a lower bound on the precision of the characteristic polynomial.
This polygon lies above the Hodge polygon of the matrix;
we give statistics on the difference and the amount of diffuse precision.
Finally, we apply Proposition \ref{prop:precision} to LU factorization and
describe experiments with lattice-based precision analysis.

@@

\noindent
{\bf Notation.}
Throughout the paper, $K$ will refer to a complete discrete 
valuation field. Usual examples are finite extensions of $\Qp$ and 
Laurent series fields over a field. We denote by $\val : K \twoheadrightarrow \Z \cup 
\{+\infty\}$ the valuation on $K$, by 
$\OK$ the ring of integers and by $\pi \in K$ an element of valuation 
$1$. We let $\Vert \cdot \Vert$ be the norm associated to $\val$.

@@

In \cite{caruso-roe-vaccon:14a}, we suggest the 
use of lattices to represent the precision of elements in 
$K$-vector spaces.  We shall contrast this approach with the
\emph{coordinate-wise method} used in Sage and Magma,
where the precision of an element is specified by giving the precision
of each coordinate separately.

Consider a finite 
dimensional\footnote{The framework of \cite{caruso-roe-vaccon:14a} is 
actually those of Banach spaces. However, we will not need infinite 
dimensional spaces here.} normed vector space $E$ 
defined over $K$. We use the notation $\Vert \cdot \Vert_E$ for the norm 
on $E$ and $B^-_E(r)$ (resp. $B^{\phantom -}_E(r)$) for the open (resp. 
closed) ball of radius $r$ centered at the origin. A \emph{lattice} $L \subset 
E$ is a sub-$\O_K$-module which generates $E$ over $K$. 
Since we are working in a ultrametric world, the balls $B^{\phantom 
-}_E(r)$ and $B^-_E(r)$ are examples of lattices. Actually, lattices 
should be thought of as special neighborhoods of $0$ and therefore are good 
candidates to model precision data. Moreover, as revealed in 
\cite{caruso-roe-vaccon:14a}, they behave quite well under (strictly) 
differentiable maps:

@@

This proposition enables the \emph{lattice method} of tracking precision,
where the precision of the input is specified as a lattice $H$ and precision
is tracked via differentials of the steps within a given algorithm.
The equality sign in Eq.~\eqref{eq:firstorder} shows that this method
yields the optimum possible precision.

In \cite{caruso-roe-vaccon:14a}, we also explained that if 
$f$ is locally analytic, then the constant $\delta$ appearing in
Proposition \ref{prop:precision} can be expressed in terms of the
growing function $\Lambda(f)$ defined by
\[
\textstyle \Lambda(f)(v) = 
\log \big( \sup_{h \in B^-_E(e^v)} \Vert 
f(h) \Vert \big)
\]
with the convention that $\Lambda(f)(v) = +\infty$ if $f$ does not
converge on $B^-_E(e^v)$.
We refer to \cite[Proposition 3.12]{caruso-roe-vaccon:14a} for the
precise statement.  We state here the case of
integral polynomial functions. A function $f : E \to F$ is said to be
\emph{integral polynomial} if it is given by multivariate polynomial 
functions with coefficients in $\O_K$ in any (equivalently all) system 
of coordinates associated to a $\OK$-basis of $B_E(1)$.

@@

In \cite[Appendix A]{caruso-roe-vaccon:14a}, the theory is extended 
to manifolds over $K$, where the precision 
datum at some point $x$ is a lattice in the tangent space at $x$. 
Propositions \ref{prop:precision} and \ref{prop:precision2} have
analogues obtained by working in charts.  We use this extension
to compute with vector spaces in \S \ref{sec:vectorspaces}.

We will use the following definition in contrasting lattice and coordinate-wise methods. 
Suppose $E$ is equipped with a basis $(e_1, \ldots, e_n)$ and write $\pi_i : E \to Ke_i$ for the projections.

\begin{deftn} \label{def:diffused}
If $H \subset E$ is a lattice, let
\[
H_0 = \pi_1(H) \oplus \cdots \oplus \pi_n(H).
\]
Set $k = \text{length}(H_0/H)$.  Then we say that $H$ has $k$
\emph{diffused digits of precision}.
\end{deftn}
If $H$ represents the actual precision of some object, then
$H_0$ is the smallest diagonal lattice containing $H$.  Since
coordinate-wise methods cannot yield a precision better than $H_0$,
$k$ provides a lower bound on the number of $p$-adic digits
gained by lattice methods over standard methods.

@@

We assume that we are given in addition two nondecreasing convex 
functions $\Lambda_g$ and $\Lambda_h$ such that $\Lambda(g) \leq 
\Lambda_g$ and $\Lambda(h) \leq \Lambda_h$. We suppose further that 
there exists $\nu$ such that $\Lambda_g$ is constant on the interval 
$({-}\infty, \nu]$\footnote{We note that this assumption is fulfilled if 
we take $\Lambda_g = \Lambda(g)$ because we have assumed that $g(0)$ 
does not vanish.}. We introduce the functions $\tau_\nu$ and $\Lambda_f$ 
defined by:
\begin{align*}
\tau_\nu(x) &= \smash{\left\{\begin{array}{ll} x & \mbox{ if } x \le \nu, \\
+\infty & \mbox{ otherwise;} \end{array}\right.} \\[8pt]
\mbox{and } \Lambda_f(x) &= \tau_\nu \circ (\id + \Lambda_g \circ \Lambda_h)(x + \alpha),
\end{align*}
%$$\begin{array}{rll}
%&\tau_\nu(x) = x & \text{if } x \leq \nu \\
%& \hphantom{\tau_\nu(x)} = {+} \infty & \text{otherwise} \medskip \\
%\text{and} &
%\multicolumn{2}{l}{\Lambda_f(x) = 
%  \tau_\nu \circ (\id + \Lambda_g \circ \Lambda_h)(x + \alpha)}
%\end{array}$$
where $\alpha$ is a real number satisfying $\Vert n! \Vert \geq 
e^{-\alpha n}$ for all $n$. If $p$ is the characteristic of the residue field,
a suitable value for $\alpha$ is $\alpha = - 
\frac p {p-1} \cdot \log \Vert p \Vert$ if $p > 0$ and $\alpha = 0$ if $p = 0$.
The next Proposition is proved in Appendix \ref{app:proof}.

@@

Note that multiplication of many random matrices plays a central
role in the theory of random walks on homogeneous spaces \cite{Benoist:2012}.
Better stability in computing such products helps estimate
Lyapunov exponents in that context.

@@

The computation of the differential of $\det : M_{n,n}(K) \to K$
is classical: at a point $M$ it is the linear map
\[
\detp(M) : dM \mapsto \tr(\com(M) \cdot dM),
\]
where $\com(M)$ stands for the comatrix of $M$, which is $\det(M) M^{-1}$ when $M$ is invertible. If $\rank(M) < n - 1$,
then $\detp(M)$ is not surjective and we cannot apply Proposition 
\ref{prop:precision}. Therefore, we suppose that $\rank(M) \ge n-1$ for the rest of this section.

@@

Recall that the \emph{Newton polygon} $\NP(f)$ of a polynomial $f(X) = 
\sum_{i} a_i X^i$ is the lower convex hull of the points $(i, 
\val(a_i))$ and the \emph{Newton polygon} $\NP(M)$ of a matrix $M$ is 
$\NP(\charpoly(M))$.  The \emph{Hodge polygon} $\HP(M)$ of $M$ is the lower 
convex hull of the points $(i, \sum_{j = 1}^{n-i} \sigma_j(M))$. For 
any matrix $M$, the polygon $\NP(M)$ lies above $\HP(M)$ 
\cite{kedlaya:padicDiffEq}*{Thm. 4.3.11}.

Such polygons arise naturally in tracking the precision of polynomials 
\cite{caruso-roe-vaccon:14a}*{\S 4.2}.  Any such polygon $P$ yields a 
lattice $\mathcal{L}_P$ in $K_{<n}[X]$ consisting of polynomials whose 
Newton polygons lie above $P$.  This lattice is generated by monomials 
$a_iX^i$, where $\val(a_i)$ is the ceiling of the height of $P$ at $i$.  
These polygons are used in a coordinate-wise precision tracking for polynomial
arithmetic. We now introduce another polygon, bounded 
between $\NP(M)$ and $\HP(M)$, that will provide an estimate on the precision of $\charpoly(M)$.

@@

It is clear from the definition $\mathcal L_{\PP(M)} \subset 
\charp(M)(\mathcal{L}_0)$ where $\mathcal{L}_0$ is the standard lattice 
$M_{n,n}(\OK)$. More precisely, $\PP(M)$ is the smallest polygon $P$ for 
which the inclusion $\mathcal L_P \subset \charp(M)(\mathcal{L}_0)$ holds. By 
Proposition \ref{prop:precision}, the precision polygon determines the minimal 
precision losses possible when encoding polynomial precision using polygons.

@@

It remains to prove the comparison with the Newton polygon.
Set $f = \charpoly(M)$, set $m_{i,j}$ as the $(i,j)$-th entry of $M$, 
$f_{i,j}$ as the $(i,j)$-th entry of $\com(X{-}M)$ and $\mu_{i,j} = 
\val(m_{i,j})$. We write $f[k]$ 
for valuation of the coefficient of $X^k$ in $f$, and set $f[-1] = 
+\infty$. The equation $(X{-}M) \cdot \com(X{-}M) = f \cdot I$ yields, 
for all $i$ and $k$,
\begin{align*}
f[k] &\ge \inf(f_{i,i}[k{-}1], \mu_{i,0} + f_{0,i}[k], \ldots, 
\mu_{i,n} + f_{n,i}[k]) \\
&\ge \inf(f_{i,i}[k{-}1], f_{j,i}[k]),
\end{align*}
with the infimum over $j$. 
Taking lower convex hulls and noting that $\PP(M)$ is nonincreasing,
which follows from the comparison with the Hodge polygon, we get the 
result.
\end{proof}

@@

For randomly chosen matrices, approximating $\mathcal{E}$ 
using the Hodge polygon loses only a small amount of precision. However,
if the $\sigma_i(M)$'s are large or if $M$ is a translate of
such a matrix, using lattice precision can be very useful.

@@

%This example has been previously studied by Caruso in \cite{caruso:12a}, 
%but without using differential precision. We are going to show here that 
%the framework of \S \ref{ssec:boundLambdaf} applies directly.
In this section, we denote by $\Vert \cdot \Vert$ the subordinate matrix norm on $M_n(K)$
and, given a positive real number $C$, we let $B(C)$ be the closed ball 
in $M_n(K)$ centered at the origin of radius $C$. We consider
the following subsets of $M_n(K)$:

@@

\noindent
{\bf Numerical experiments.}
Let $B_n=(E_{i,j})_{1 \leq i,j\leq n}$ be the canonical basis of 
$M_n(K)$. It can be naturally seen as a basis of $L_n^0 \times U_n$ as
well. For a given $M \in O_n$, let us abuse notation and write $D'(M)$
for the matrix of this linear mapping in the above basis. We remark that
$D'(M)$ is lower-triangular in this basis. Projecting $D'(M)$ onto each
coordinate, we find the best coordinate-wise loss of precision we can 
hope for the computation of $D$ is given by $\sum_u \left( \max_v \left( 
\val(D'(M)_{u,v}) \right) \right)$. This number should be compared to
$\val(\det(D'(M)))$, which is precision lost in the lattice method. 
%With the language of Definition \ref{def:diffused},
The number of diffused digits of precision of $D'(M)(M_{n,n}(\OK))$ is 
then the difference between these two numbers.
Figure \ref{fig:LU} summarizes the mean and standard deviation of those 
losses for a sample of 2000 random matrices in $M_{d,d}(\mathbb{Z}_2)$.
%
\begin{figure}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|c|c|c|c|c|}
\hline 
& \multicolumn{4}{|c|}{Loss of precision in LU decomposition} \\
\cline{2-5}
\raisebox{0.2em}{matrix}
& \multicolumn{2}{|c|}{coord-wise method} & \multicolumn{2}{|c|}{lattice method}  \\  \cline{2-5}
\smash{\raisebox{0.6em}{size}} 
& \hspace{0.5em}mean\hspace{0.5em} & deviation 
& \hspace{0.5em}mean\hspace{0.5em} & deviation \\ \hline
$2$ & $\phantom{2}3.0$& $5$ & $1.5$ & $1.4$ \\% \hline
$3$& $\phantom{2}9.4$& $11$ & $2.3$ & $2.3$\\% \hline
$4$ & $20\phantom{.4}$& $20$ & $3.8$ & $3.1$\\ \hline
\end{tabular} 
\smallskip

@@

\noindent
{\bf Inverse images.}
We now consider the inverse image mapping $\II$ sending a 
pair $(f,W) \in \mathcal W = \Hom(E,F) \times \Grass(F,d)$ to 
$f^{-1}(W)$. As before, this map does not take values in a single 
Grassmannian, so we need to stratify $\mathcal W$ in order to get 
differentiable functions. For each integer $s \in [0,n]$, we introduce 
the submanifold $\mathcal W_s$ of $\mathcal W$ consisting of those 
pairs $(f,W)$ such that $\dim f^{-1}(W) = s$. For all $s$, $\II$ 
induces a continuous function
$\II_s : \mathcal W_s \to \Grass(E,s)$.
Pick $(f,W) \in \mathcal W_s$. Set $V = f^{-1}(W)$ and denote by $w : F 
\to F/W$ the canonical projection.
Similarly to what we have done for direct images, one can prove that
the tangent space of $\mathcal W_s$ at some point $(f,W) \in \mathcal
W_s$ is the subspace of $\Hom(E,F) \times \Hom(W,F/W)$ consisting of 
pairs $(df,dw)$ such that
$d \tilde v = (w \circ df + dw \circ f)_{|W}$
factors through the linear mapping $\varphi : E/V \to F/W$ induced by 
$f$. Furthermore $\II_s$ is differentiable at $(f,W)$ and its 
differential is the linear mapping that takes $(df,dw)$ as above to the 
unique element $dv \in \Hom(V,E/V)$ satisfying $\varphi \circ dv =
d\tilde v$.

Direct images and inverse images are related by duality
as follows: if $f : E \to F$ is any linear map and $W$ is a subspace
of $F$, then $f^\star(W^\perp) = f^{-1}(W)^\perp$. We 
can thus deduce the differentials of $\DI_s$ from those of $\II_s$ and 
\emph{vice versa}.

@@

\noindent
{\bf Sums and intersections.}
Let $d_1$ and $d_2$ be two nonnegative integers. We consider the 
function $\Sigma$ defined on the manifold $\mathcal C = \Grass(E,d_1) 
\times \Grass(E,d_2)$ by $\Sigma(V_1, V_2) = V_1 + V_2$. As before, in 
order to study $\Sigma$, we stratify $\mathcal C$ according to the 
dimension of the sum: for each integer $d \in [0, d_1+d_2]$, we define 
$\mathcal C_d$ as the submanifold of $\mathcal C$ consisting of those 
pairs $(V_1, V_2)$ such that $\dim(V_1 + V_2) = d$. We get a 
well-defined mapping $\mathcal C_d \to \Grass(E,d)$ whose differential
can be computed as before. The
tangent space of $\mathcal C_d$ at a given point $(V_1, V_2)$ consists 
of pairs $(dv_1, dv_2) \in \Hom(V_1, E/V_1)
\times \Hom(V_2, E/V_2)$ such that $dv_1 \equiv dv_2 \pmod{V_1 + V_2}$ 
on the intersection $V_1 \cap V_2$, and the differential of $\Sigma$
at $(V_1, V_2)$ maps $(dv_1, dv_2)$ to $dv \in \Hom(V, E/V)$ (with $V
= V_1 + V_2$) defined by $dv(v_1 + v_2) = dv_1(v_1) + dv_2(v_2)$ ($v_1
\in V_1$, $v_2 \in V_2$).
